---
title: "Zebrafish Data Processing and Mergin"
author: "Oliver F. Anderson"
date: "2023-10-27"
output: html_document
---

**Zebrafish Data Merging and Wrangling Documentation**
Initialize the work environment.
```{r}
library(tidyverse)
```

Next process the data from PsicQuic into a nice format.
```{r}
# get interactor data from PsicQuic
psicquic_interactors <- read_csv("interactors.csv")

# how many unique interactors? not very many...
length(unique(psicquic_interactors$ID))

# make sure that there are no missing values
psicquic_interactors <- subset(psicquic_interactors, !is.na(ID))

psicquic_interactors <- psicquic_interactors %>% distinct(ID, .keep_all = TRUE)

psicquic_interactors <- psicquic_interactors[,c(1,2,13,15,18,19)]

colnames(psicquic_interactors) <- c("ID", "shortLabel",
                                    "primaryRefDb", "primaryRefId",
                                    "secondaryRefDb", "secondaryRefId")

# it appears we are missing two values, let's manually find these since there are only 2
# subset(psicquic_interactors, is.na(uniprotId))
psicquic_interactors$uniprotId <- case_when(
    psicquic_interactors$primaryRefDb == "uniprotkb" ~ psicquic_interactors$primaryRefId,
    psicquic_interactors$secondaryRefDb == "uniprotkb" ~ psicquic_interactors$secondaryRefId,
    psicquic_interactors$shortLabel == "EBI-2213356" ~ "Q2YDS1",
    psicquic_interactors$shortLabel == "EBI-1536909" ~ "A2A121"
  )


psicquic_interactorsIDs <- psicquic_interactors %>% select(ID, uniprotId)

# get interaction data from PsicQuic
psicquic_interactions <- read_csv("interactions.csv")
# change names to more readable ones
colnames(psicquic_interactions) <- c("ID", "InteractorRef1", "InteractorRef2")

# first join the IDs for Ref1
psicquic_interactome <- left_join(psicquic_interactions, psicquic_interactorsIDs, by = c("InteractorRef1" = "ID"), relationship="many-to-many")

# then join the IDs for Ref2
psicquic_interactome <- left_join(psicquic_interactome, psicquic_interactorsIDs, by = c("InteractorRef2" = "ID"), relationship="many-to-many")

colnames(psicquic_interactome) <- c("interactionID", "InteractorRef1", "InteractorRef2", 'uniprotID1', "uniprotID2")

# extract only the necessary information
psicquic_interactome <- 
  psicquic_interactome %>% 
  select(uniprotID1, uniprotID2) %>%
  mutate(source = "psicquic")

psicquic_interactome <- psicquic_interactome %>% 
  mutate(evidence = "experimental")
```

Save the PsicQuic data to a separate file
```{r}
write_csv(psicquic_interactome, "./zfish_psicquic_results.csv")
```

This is the less stringent version for data mining, filtered by experimental, database, and textmined data. Resulting interactome had 11766 unique proteins.
```{r}
# read the file in
string_db_interactome2 <- read_delim("7955.protein.physical.links.full.v12.0_attempt2.txt", delim = " ")

# get only the data from String-DB that has experimental evidence
string_db_interactome2 <- string_db_interactome2 %>% filter(experiments > 0 |
                                                            database > 0 |
                                                            textmining > 0)
string_db_interactome2 <- string_db_interactome2 %>% 
  mutate(
      evidence = case_when(
        experiments > 0 ~ "experimental",
        (database > 0 & experiments <= 0) ~ "database",
        (textmining > 0 & experiments <= 0 & database <= 0) ~ "textmined")
      )


# select only the relevant information
string_db_interactome2 <- string_db_interactome2 %>% select(protein1, protein2, evidence)

# get the mapping data to get uniprot aliases
string_db_names2 <- read_tsv("7955.protein.aliases.v12.0.txt")

string_db_names2 <- string_db_names2 %>%
  filter(source == "UniProt_AC") %>% 
  select(string_protein_id, alias)

# join the uniprot IDs for protein1
string_db_interactome2 <- left_join(string_db_interactome2, string_db_names2, by = c("protein1" = "string_protein_id"), relationship = "many-to-many")

# join the uniprot IDs for protein2
string_db_interactome2 <- left_join(string_db_interactome2, string_db_names2, by = c("protein2" = "string_protein_id"), relationship = "many-to-many")

# rename the columns to prepare to merge datasets
colnames(string_db_interactome2) <- c("protein1", "protein2", "evidence", "uniprotID1", "uniprotID2")

# extract only the relevant information
string_db_interactome2 <- 
  string_db_interactome2 %>%
  select(uniprotID1, uniprotID2, evidence) %>%
  mutate(source = "string-db")

View(string_db_interactome2)
```


Now, merge the PsicQuic and String-DB interactomes together.
```{r}
zfish_interactome <- rbind(string_db_interactome2, psicquic_interactome)
```

```{r}
write_tsv(zfish_interactome, "./temp_updated_Mar12_24_zfish_interactome.txt")
```

Get list of unique proteins.
```{r}
uniprotIDs <- unique(c(zfish_interactome$uniprotID1, zfish_interactome$uniprotID2))


write_csv(as_tibble(uniprotIDs), "./updated_Mar12_24_zfish_legacy_uniprotIDs.txt")
```

Check lengths to see number of unique nodes.
```{r}
length(uniprotIDs)
```

`From` column is the initial Unique ID while `Entry` column is what UniProt currently stores the protein as, thus we need to convert to the current entries in our interactome. This will reduce the number of unique entries by combining them into the official UniProt accession IDs.
```{r}
common_names2 <- read_tsv("zfish_id_mapper2.tsv")

# From column is the initial Unique ID while Entry column is what UniProt currently stores the entry as, thus we need to convert to the current entries in our interactome

# Merge the two datasets based on the matching columns
temp_df <- inner_join(zfish_interactome, common_names2, by=c("uniprotID1" = "From"))

temp_df <- temp_df %>% select(c(Entry, uniprotID2, source, evidence))
colnames(temp_df) <- c("uniprotID1", "uniprotID2", "source", "evidence")

temp_df <- inner_join(temp_df, common_names2, by=c("uniprotID2" = "From"))
zfish_interactome <- temp_df %>% select(c(uniprotID1, Entry, source, evidence))
colnames(zfish_interactome) <- c("uniprotID1", "uniprotID2", "source", "evidence")

View(zfish_interactome)
```

Once we have converted to the proper UniprotID, we need to now import the common names.
```{r}
common_names2 <- common_names2 %>% select(Entry, `Protein names`)

zfish_interactome <- inner_join(zfish_interactome, common_names2, by = c("uniprotID1" = "Entry"))

zfish_interactome <- inner_join(zfish_interactome, common_names2, by = c("uniprotID2" = "Entry"))

colnames(zfish_interactome) <- c("uniprotID1", "uniprotID2", "source", "evidence", "alt_name1", "alt_name2")

View(zfish_interactome)
```

Many entries have "deleted" as their common name value, if we want to remove these the following code will do so:
```{r}
deleted_interactions <- zfish_interactome %>% filter(alt_name1 == "deleted") %>% filter(alt_name2 == "deleted")

zfish_interactome_deleted <- zfish_interactome %>% filter(alt_name1 != "deleted") %>% filter(alt_name2 != "deleted")

zfish_interactome_deleted <- zfish_interactome_deleted %>% mutate(name1 = str_extract(alt_name1, "^[^(]*"),
                             name2 = str_extract(alt_name2, "^[^(]*"))

View(zfish_interactome_deleted)
# number of deleted proteins
print("Number of deleted proteins:")
length(unique(c(deleted_interactions$uniprotID1, deleted_interactions$uniprotID2)))

# number of unique proteins left over
print("Number of kept proteins:")
uniqueProteins <- unique(c(zfish_interactome_deleted$uniprotID1, zfish_interactome_deleted$uniprotID2))
length(uniqueProteins)
```
6,438 nodes is more data than the previous method!

Now there appears to be a problem with the naming convention, thus we will add gene_name as a column to use for the unique name. 
```{r}
gene_names <- read_tsv("zfish_gene_names.tsv") %>% select(From, `Protein names`, Gene_Names_primary)

length(unique(gene_names$From))
length(unique(gene_names$Gene_Names_primary))
sum(is.na(gene_names$Gene_Names_primary))
sum(is.na(gene_names$`Protein names`))
sum(is.na(gene_names$From))

zfish_interactome_deleted <- zfish_interactome_deleted %>% select(uniprotID1, uniprotID2, source, evidence)


zfish_interactome_deleted <- inner_join(zfish_interactome_deleted, gene_names, by = c("uniprotID1" = "From"))

zfish_interactome_deleted <- inner_join(zfish_interactome_deleted, gene_names, by = c("uniprotID2" = "From"))

colnames(zfish_interactome_deleted) <- c("uniprotID1", "uniprotID2", "source", "evidence", "alt_name1", "gene_name1", "alt_name2", "gene_name2")

zfish_interactome_deleted <- zfish_interactome_deleted %>% mutate(
  name1 = case_when(
    !is.na(gene_name1) ~ gene_name1,
    is.na(gene_name1) ~ alt_name1
  ),
  name2 = case_when(
    !is.na(gene_name2) ~ gene_name2,
    is.na(gene_name2) ~ alt_name2
  )
)

zfish_interactome_deleted$name1 <- gsub("\\(.*", "", zfish_interactome_deleted$name1)

zfish_interactome_deleted$name2 <-  gsub("\\(.*", "", zfish_interactome_deleted$name2)
```


```{r}
write_tsv(zfish_interactome_deleted, "./updated_Mar12_24_zfish_interactome.txt")
write_tsv(as_tibble(uniqueProteins), "./updated_Mar12_24_zfish_unique_protein_ids.txt")
View(zfish_interactome_deleted)
```

```{r}
# these are the "Reviewed" annotations downloaded from https://www.ebi.ac.uk/QuickGO/annotations?taxonId=7955&taxonUsage=descendants&geneProductSubset=Swiss-Prot&geneProductType=protein
go_terms <- read_tsv("zfish_GO_annotations_Mar12_24.tsv")
colnames(go_terms) <- gsub(" ", "_", colnames(go_terms))
write_tsv(go_terms, "zfish_GO_data_Mar12_24.tsv")
```



```{r}
setwd("~/Desktop/GitHub_Repos/bio-net-viz/data/DanioRerio/")
df <- read_tsv("~/Desktop/GitHub_Repos/bio-net-viz/data/DanioRerio/updated_Mar12_24_zfish_interactome.txt")
View(df)

df %>% 
  group_by(alt_name1) %>% 
  View()

df$alt_name1 <- df$alt_name1 %>% str_trim()
df$alt_name2 <- df$alt_name2 %>% str_trim()
df$name1 <- df$name1 %>% str_trim()
df$name2 <- df$name2 %>% str_trim()

write_tsv(df, "zfish_interactome_Mar12_2024.txt")
```


