---
title: "B. Subtilis Data Processing and Merging"
author: "Oliver F. Anderson"
date: "2024-05-31"
output: html_document
---

**B. Subtilis Data Merging and Wrangling Documentation**
Initialize the work environment.
```{r}
library(tidyverse)
```

This is non-stringent data mining, filtered by experimental, database, and textmined data. The only thing we filter out is homology-inferred data.
```{r}
setwd("~/Desktop/GitHub_Repos/bio-net-viz/data/BacillusSubtilis/")
# read the file in
string_db_interactome <- read_delim("224308.protein.physical.links.full.v12.0.txt", delim = " ")

# get only the data from String-DB that has experimental evidence
string_db_interactome <- string_db_interactome %>% filter(experiments > 0 |
                                                            database > 0 |
                                                            textmining > 0)
string_db_interactome <- string_db_interactome %>% 
  mutate(
      evidence = case_when(
        experiments > 0 ~ "experimental",
        (database > 0 & experiments <= 0) ~ "database",
        (textmining > 0 & experiments <= 0 & database <= 0) ~ "textmined")
      )


# select only the relevant information
string_db_interactome <- string_db_interactome %>% select(protein1, protein2, evidence)

# get the mapping data to get uniprot aliases
string_db_names2 <- read_tsv("224308.protein.info.v12.0.txt")

string_db_names2 <- string_db_names2 %>%
  select(string_protein_id, preferred_name)

# join the uniprot IDs for protein1
string_db_interactome <- left_join(string_db_interactome, string_db_names2, by = c("protein1" = "string_protein_id"), relationship = "many-to-many")

# join the uniprot IDs for protein2
string_db_interactome <- left_join(string_db_interactome, string_db_names2, by = c("protein2" = "string_protein_id"), relationship = "many-to-many")

# rename the columns to prepare to merge datasets
colnames(string_db_interactome) <- c("protein1", "protein2", "evidence", "name1", "name2")

# extract only the relevant information
string_db_interactome <-
  string_db_interactome %>%
  select(protein1, name1, protein2, name2, evidence) %>%
  mutate(source = "string-db")

# remove txid tag from beginning of BSU_ID
string_db_interactome$protein1 <- gsub("224308.", "", string_db_interactome$protein1)

string_db_interactome$protein2 <- gsub("224308.", "", string_db_interactome$protein2)

# capitalize first letter of protein name
string_db_interactome$name1 <- gsub("\\b([a-z])", "\\U\\1", string_db_interactome$name1, perl=TRUE)

string_db_interactome$name2 <- gsub("\\b([a-z])", "\\U\\1", string_db_interactome$name2, perl=TRUE)

colnames(string_db_interactome) <- c("protein_1_locus", "protein_1_name", "protein_2_locus", "protein_2_name", "evidence", "source")

# Add an underscore to match the other BSU convention
string_db_interactome$protein_1_locus <- gsub("BSU", "BSU_", string_db_interactome$protein_1_locus)

string_db_interactome$protein_2_locus <- gsub("BSU", "BSU_", string_db_interactome$protein_2_locus)

# Number of unique proteins
length(unique(c(unique(string_db_interactome$protein_1_name), unique(string_db_interactome$protein_2_name))))

View(string_db_interactome)
```
The resulting interactome from STRING-DB had 1446 unique proteins.

Save the STRING-DB data to a separate file
```{r}
setwd("~/Desktop/GitHub_Repos/bio-net-viz/data/BacillusSubtilis/")
write_csv(string_db_interactome, "./bsub_string_db_results.csv")
```

```{r}
setwd("~/Desktop/GitHub_Repos/bio-net-viz/data/BacillusSubtilis/")
subtiwiki_interactome <- read_csv("bsub_interactome.csv")

# Add columns to match data structure of STRING DB
subtiwiki_interactome <- subtiwiki_interactome %>% mutate(
  evidence = "experimental",
  source = "subtiwiki")

# Number of unique proteins in subtiwiki interactome
length(unique(c(unique(subtiwiki_interactome$protein_1_name), unique(subtiwiki_interactome$protein_2_name))))
```


Now, merge the SubtiWiki and STRING-DB interactomes together.
```{r}
bsub_interactome <- rbind(string_db_interactome, subtiwiki_interactome)

length(unique(c(unique(bsub_interactome$protein_1_name), unique(bsub_interactome$protein_2_name))))

# During import we noticed some IDs have duplicate names. Let's fix this and set an alt_name parameter.
protein_1_duplicates <- bsub_interactome %>%
  group_by(protein_1_locus) %>%
  filter(n_distinct(protein_1_name) > 1) %>%
  arrange(protein_1_locus)

protein_2_duplicates <- bsub_interactome %>%
  group_by(protein_2_locus) %>%
  filter(n_distinct(protein_2_name) > 1) %>%
  arrange(protein_2_locus)

duplicates <- rbind(protein_1_duplicates, protein_2_duplicates)

length(unique(c(unique(duplicates$protein_1_locus), unique(duplicates$protein_2_locus))))
```

Sub the blank space in column names with underscore for Neo4j import.
```{r}
# these are the "Reviewed" annotations downloaded from https://www.ebi.ac.uk/QuickGO/annotations?taxonId=7955&taxonUsage=descendants&geneProductSubset=Swiss-Prot&geneProductType=protein
setwd("~/Desktop/GitHub_Repos/bio-net-viz/data/BacillusSubtilis/")

go_terms <- read_tsv("bsub_GO_annotations_2024-06-03.txt")
colnames(go_terms) <- gsub(" ", "_", colnames(go_terms))

go_terms$SYMBOL <- gsub("\\b([a-z])", "\\U\\1", go_terms$SYMBOL, perl=TRUE)

write_tsv(go_terms, "bsub_GO_annotations_2024-06-03.txt")
```

```{r}
setwd("~/Desktop/GitHub_Repos/bio-net-viz/data/BacillusSubtilis/")

BSU_to_Uniprot <- read_tsv("subtiwiki.gene.export.2024-06-03.tsv")

BSU_to_Uniprot <- BSU_to_Uniprot %>% 
  select(locus, GENE_PRODUCT_ID)

# join datasets together
bsub_GO_data <- inner_join(go_terms, BSU_to_Uniprot, "GENE_PRODUCT_ID", relationship="many-to-many")

# write to csv
write_tsv(bsub_GO_data, "bsub_GO_annotations_2024-06-03.txt")
```


Sometimes the names will have trailing blank spaces so let's ensure that they won't with the following code. Finally, we will write out the final zebrafish interactome.
```{r}
setwd("~/Desktop/GitHub_Repos/bio-net-viz/data/BacillusSubtilis/")
write_tsv(bsub_interactome, "bsub_interactome_May31_2024.txt")
```